---
title: "Comparison: 2023 vs 2025"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparison: 2023 vs 2025}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(PERiskProfileR)
library("data.table")
library("ggplot2")
theme_set(theme_bw())

df_23 <- get_validation_data(model = "FMF2023")
df_25 <- get_validation_data(model = "FMF2025")

common_ids <- intersect(df_23$id, df_25$id)
df_23 <- df_23[id %in% common_ids]
df_25 <- df_25[id %in% common_ids]

common_names <- intersect(names(df_23), names(df_25))
identical_cols <- Filter(function(col) {
  isTRUE(all.equal(df_23[[col]], df_25[[col]]))
}, common_names)

df <- merge(df_23, df_25, by = identical_cols, suffixes = c(".x", ".y"))
```

---

## Part 1: Multiple of the Median (MoM) Validation

### 1.1 MoM MAP

**Mean Arterial Pressure** is a key biophysical marker adjusted for maternal characteristics.

#### Agreement Plot

```{r mom_map_agreement, fig.height=5, fig.width=7}
# Filter out missing values
df_map <- df[!is.na(mom_map_analytical.x) & !is.na(mom_map_analytical.y)]

ggplot(df_map, aes(x = mom_map_analytical.x, y = mom_map_analytical.y)) +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed", linewidth = 1) +
  geom_point(alpha = 0.3, size = 1.5) +
  coord_equal(xlim = c(0.5, 1.5), ylim = c(0.5, 1.5)) +
  labs(
    title = "MoM MAP: FMF-2023 vs FMF-2025",
    x = "FMF-2023",
    y = "FMF-2025",
    caption = "Red line: perfect agreement (y = x)"
  ) +
  theme_bw()
```

#### Bland-Altman Plot

```{r mom_map_bland_altman, fig.height=5, fig.width=7}
df_map[, mean_map := (mom_map_analytical.x + mom_map_analytical.y) / 2]
df_map[, diff_map := mom_map_analytical.y - mom_map_analytical.x]

mean_diff <- mean(df_map$diff_map, na.rm = TRUE)
sd_diff <- sd(df_map$diff_map, na.rm = TRUE)
loa_upper <- mean_diff + 1.96 * sd_diff
loa_lower <- mean_diff - 1.96 * sd_diff

ggplot(df_map, aes(x = mean_map, y = diff_map)) +
  geom_hline(yintercept = 0, color = "black", linetype = "solid") +
  geom_hline(yintercept = mean_diff, color = "blue", linetype = "dashed", linewidth = 0.8) +
  geom_hline(yintercept = loa_upper, color = "red", linetype = "dotted", linewidth = 0.8) +
  geom_hline(yintercept = loa_lower, color = "red", linetype = "dotted", linewidth = 0.8) +
  geom_point(alpha = 0.3, size = 1.5) +
  labs(
    title = "Bland-Altman Plot: MoM MAP",
    x = "Mean of FMF-2023 and FMF-2025",
    y = "Difference (FMF-2025 - FMF-2023)",
    caption = sprintf("Mean difference: %.4f | 95%% LoA: [%.4f, %.4f]", 
                      mean_diff, loa_lower, loa_upper)
  ) +
  theme_bw()
```

#### Summary Statistics

```{r mom_map_stats}
cat("MoM MAP Agreement Statistics:\n")
cat(sprintf("  Mean difference: %.6f\n", mean_diff))
cat(sprintf("  SD of differences: %.6f\n", sd_diff))
cat(sprintf("  95%% Limits of Agreement: [%.6f, %.6f]\n", loa_lower, loa_upper))
cat(sprintf("  Correlation: %.4f\n", cor(df_map$mom_map_analytical.x, df_map$mom_map_analytical.y)))
```

---

### 1.2 MoM UtPI

**Uterine Artery Pulsatility Index** reflects placental resistance.

#### Agreement Plot

```{r mom_utpi_agreement, fig.height=5, fig.width=7}
df_utpi <- df[!is.na(mom_utpi_analytical.x) & !is.na(mom_utpi_analytical.y)]

ggplot(df_utpi, aes(x = mom_utpi_analytical.x, y = mom_utpi_analytical.y)) +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed", linewidth = 1) +
  geom_point(alpha = 0.3, size = 1.5) +
  coord_equal(xlim = c(0.3, 2.0), ylim = c(0.3, 2.0)) +
  labs(
    title = "MoM UtPI: FMF-2023 vs FMF-2025",
    x = "FMF-2023",
    y = "FMF-2025",
    caption = "Red line: perfect agreement (y = x)"
  ) +
  theme_bw()
```

#### Bland-Altman Plot

```{r mom_utpi_bland_altman, fig.height=5, fig.width=7}
df_utpi[, mean_utpi := (mom_utpi_analytical.x + mom_utpi_analytical.y) / 2]
df_utpi[, diff_utpi := mom_utpi_analytical.y - mom_utpi_analytical.x]

mean_diff_utpi <- mean(df_utpi$diff_utpi, na.rm = TRUE)
sd_diff_utpi <- sd(df_utpi$diff_utpi, na.rm = TRUE)
loa_upper_utpi <- mean_diff_utpi + 1.96 * sd_diff_utpi
loa_lower_utpi <- mean_diff_utpi - 1.96 * sd_diff_utpi

ggplot(df_utpi, aes(x = mean_utpi, y = diff_utpi)) +
  geom_hline(yintercept = 0, color = "black", linetype = "solid") +
  geom_hline(yintercept = mean_diff_utpi, color = "blue", linetype = "dashed", linewidth = 0.8) +
  geom_hline(yintercept = loa_upper_utpi, color = "red", linetype = "dotted", linewidth = 0.8) +
  geom_hline(yintercept = loa_lower_utpi, color = "red", linetype = "dotted", linewidth = 0.8) +
  geom_point(alpha = 0.3, size = 1.5) +
  labs(
    title = "Bland-Altman Plot: MoM UtPI",
    x = "Mean of FMF-2023 and FMF-2025",
    y = "Difference (FMF-2025 - FMF-2023)",
    caption = sprintf("Mean difference: %.4f | 95%% LoA: [%.4f, %.4f]", 
                      mean_diff_utpi, loa_lower_utpi, loa_upper_utpi)
  ) +
  theme_bw()
```

#### Summary Statistics

```{r mom_utpi_stats}
cat("MoM UtPI Agreement Statistics:\n")
cat(sprintf("  Mean difference: %.6f\n", mean_diff_utpi))
cat(sprintf("  SD of differences: %.6f\n", sd_diff_utpi))
cat(sprintf("  95%% Limits of Agreement: [%.6f, %.6f]\n", loa_lower_utpi, loa_upper_utpi))
cat(sprintf("  Correlation: %.4f\n", cor(df_utpi$mom_utpi_analytical.x, df_utpi$mom_utpi_analytical.y)))
```

---

### 1.3 MoM PlGF

**Placental Growth Factor** is a biochemical marker that varies by assay platform.

#### Agreement Plot

```{r mom_plgf_agreement, fig.height=5, fig.width=7}
df_plgf <- df[!is.na(mom_plgf_analytical.x) & !is.na(mom_plgf_analytical.y)]

ggplot(df_plgf, aes(x = mom_plgf_analytical.x, y = mom_plgf_analytical.y)) +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed", linewidth = 1) +
  geom_point(alpha = 0.3, size = 1.5) +
  coord_equal(xlim = c(0.2, 2.5), ylim = c(0.2, 2.5)) +
  labs(
    title = "MoM PlGF: FMF-2023 vs FMF-2025",
    x = "FMF-2023",
    y = "FMF-2025",
    caption = "Red line: perfect agreement (y = x)"
  ) +
  theme_bw()
```

#### Bland-Altman Plot

```{r mom_plgf_bland_altman, fig.height=5, fig.width=7}
df_plgf[, mean_plgf := (mom_plgf_analytical.x + mom_plgf_analytical.y) / 2]
df_plgf[, diff_plgf := mom_plgf_analytical.y - mom_plgf_analytical.x]

mean_diff_plgf <- mean(df_plgf$diff_plgf, na.rm = TRUE)
sd_diff_plgf <- sd(df_plgf$diff_plgf, na.rm = TRUE)
loa_upper_plgf <- mean_diff_plgf + 1.96 * sd_diff_plgf
loa_lower_plgf <- mean_diff_plgf - 1.96 * sd_diff_plgf

ggplot(df_plgf, aes(x = mean_plgf, y = diff_plgf)) +
  geom_hline(yintercept = 0, color = "black", linetype = "solid") +
  geom_hline(yintercept = mean_diff_plgf, color = "blue", linetype = "dashed", linewidth = 0.8) +
  geom_hline(yintercept = loa_upper_plgf, color = "red", linetype = "dotted", linewidth = 0.8) +
  geom_hline(yintercept = loa_lower_plgf, color = "red", linetype = "dotted", linewidth = 0.8) +
  geom_point(alpha = 0.3, size = 1.5) +
  labs(
    title = "Bland-Altman Plot: MoM PlGF",
    x = "Mean of FMF-2023 and FMF-2025",
    y = "Difference (FMF-2025 - FMF-2023)",
    caption = sprintf("Mean difference: %.4f | 95%% LoA: [%.4f, %.4f]", 
                      mean_diff_plgf, loa_lower_plgf, loa_upper_plgf)
  ) +
  theme_bw()
```

#### Summary Statistics

```{r mom_plgf_stats}
cat("MoM PlGF Agreement Statistics:\n")
cat(sprintf("  Mean difference: %.6f\n", mean_diff_plgf))
cat(sprintf("  SD of differences: %.6f\n", sd_diff_plgf))
cat(sprintf("  95%% Limits of Agreement: [%.6f, %.6f]\n", loa_lower_plgf, loa_upper_plgf))
cat(sprintf("  Correlation: %.4f\n", cor(df_plgf$mom_plgf_analytical.x, df_plgf$mom_plgf_analytical.y)))
```

---

## Part 2: Prior Risk Validation

Prior risk is based solely on maternal characteristics (history, demographics, medical conditions).

Perfect agreement (no change in model)

---

## Part 3: Final Risk Score Validation

The final risk combines prior risk with biomarker likelihoods.

### Agreement Plot (Log Scale)

```{r risk_agreement, fig.height=5, fig.width=7}
df_risk <- df[!is.na(risk_analytical.x) & !is.na(risk_analytical.y)]

ggplot(df_risk, aes(x = risk_analytical.x, y = risk_analytical.y)) +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed", linewidth = 1) +
  geom_point(alpha = 0.3, size = 1.5) +
  scale_x_log10(labels = scales::percent_format()) +
  scale_y_log10(labels = scales::percent_format()) +
  coord_equal() +
  labs(
    title = "Final Risk Score: FMF-2023 vs FMF-2025",
    x = "FMF-2023",
    y = "FMF-2025",
    caption = "Red line: perfect agreement (y = x) | Log-log scale"
  ) +
  theme_bw()
```

### Bland-Altman Plot (Log Scale)

```{r risk_bland_altman, fig.height=5, fig.width=7}
df_risk[, mean_risk := (risk_analytical.x + risk_analytical.y) / 2]
df_risk[, diff_risk := risk_analytical.y - risk_analytical.x]

mean_diff_risk <- mean(df_risk$diff_risk, na.rm = TRUE)
sd_diff_risk <- sd(df_risk$diff_risk, na.rm = TRUE)
loa_upper_risk <- mean_diff_risk + 1.96 * sd_diff_risk
loa_lower_risk <- mean_diff_risk - 1.96 * sd_diff_risk

ggplot(df_risk, aes(x = mean_risk, y = diff_risk)) +
  geom_hline(yintercept = 0, color = "black", linetype = "solid") +
  geom_hline(yintercept = mean_diff_risk, color = "blue", linetype = "dashed", linewidth = 0.8) +
  geom_hline(yintercept = loa_upper_risk, color = "red", linetype = "dotted", linewidth = 0.8) +
  geom_hline(yintercept = loa_lower_risk, color = "red", linetype = "dotted", linewidth = 0.8) +
  geom_point(alpha = 0.3, size = 1.5) +
  scale_x_log10(labels = scales::percent_format()) +
  labs(
    title = "Bland-Altman Plot: Final Risk Score",
    x = "Mean of FMF-2023 and FMF-2025 (log scale)",
    y = "Difference (FMF-2025 - FMF-2023)",
    caption = sprintf("Mean difference: %.6f | 95%% LoA: [%.6f, %.6f]", 
                      mean_diff_risk, loa_lower_risk, loa_upper_risk)
  ) +
  theme_bw()
```

### Summary Statistics

```{r risk_stats}
cat("Final Risk Agreement Statistics:\n")
cat(sprintf("  Mean difference: %.8f\n", mean_diff_risk))
cat(sprintf("  SD of differences: %.8f\n", sd_diff_risk))
cat(sprintf("  95%% Limits of Agreement: [%.8f, %.8f]\n", loa_lower_risk, loa_upper_risk))
cat(sprintf("  Correlation: %.4f\n", cor(df_risk$risk_analytical.x, df_risk$risk_analytical.y)))
cat(sprintf("  Mean absolute difference: %.8f\n", mean(abs(df_risk$diff_risk))))
```
---

## Part 4: Clinical Classification Agreement

The most critical validation is whether the analytical and online methods **agree on clinical risk categories**.

### 4.1 High Risk Threshold: 1/100 (1%)

```{r confusion_1_100}
df_risk[, high_analytical_100 := risk_analytical.x >= 0.01]
df_risk[, high_online_100 := risk_analytical.y >= 0.01]

conf_100 <- table(
  Analytical = ifelse(df_risk$high_analytical_100, "High (≥1%)", "Low (<1%)"),
  Online = ifelse(df_risk$high_online_100, "High (≥1%)", "Low (<1%)")
)

print(conf_100)

# Calculate agreement metrics
n_total <- sum(conf_100)
n_agree <- conf_100[1,1] + conf_100[2,2]
n_disagree <- conf_100[1,2] + conf_100[2,1]

accuracy <- n_agree / n_total
sensitivity <- conf_100[2,2] / (conf_100[2,2] + conf_100[1,2])
specificity <- conf_100[1,1] / (conf_100[1,1] + conf_100[2,1])
kappa <- (2 * (conf_100[1,1] * conf_100[2,2] - conf_100[1,2] * conf_100[2,1])) / 
         ((conf_100[1,1] + conf_100[1,2]) * (conf_100[1,2] + conf_100[2,2]) + 
          (conf_100[1,1] + conf_100[2,1]) * (conf_100[2,1] + conf_100[2,2]))

cat("\nClassification Agreement at 1/100 Threshold:\n")
cat(sprintf("  Total agreement: %.2f%% (%d/%d)\n", 
            100 * accuracy, n_agree, n_total))
cat(sprintf("  Disagreement: %.2f%% (%d/%d)\n", 
            100 * (1 - accuracy), n_disagree, n_total))
cat(sprintf("  Sensitivity: %.2f%%\n", 100 * sensitivity))
cat(sprintf("  Specificity: %.2f%%\n", 100 * specificity))
cat(sprintf("  Cohen's Kappa: %.4f\n", kappa))
```

```{r confusion_plot_100, fig.height=4, fig.width=6}
# Create confusion matrix plot
conf_df <- as.data.frame(conf_100)
conf_df$prop <- conf_df$Freq / sum(conf_df$Freq)

ggplot(conf_df, aes(x = Online, y = Analytical, fill = Freq)) +
  geom_tile(color = "white", linewidth = 1.5) +
  geom_text(aes(label = sprintf("%d\n(%.1f%%)", Freq, 100*prop)), 
            size = 5, fontface = "bold") +
  scale_fill_gradient(low = "white", high = "steelblue") +
  labs(
    title = "Classification Agreement: 1/100 Threshold",
    x = "Online Calculator",
    y = "Analytical Formula",
    fill = "Count"
  ) +
  theme_minimal() +
  theme(
    axis.text = element_text(size = 11, face = "bold"),
    axis.title = element_text(size = 12, face = "bold"),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5)
  )
```

---

### 4.2 High Risk Threshold: 1/150 (0.67%)

```{r confusion_1_150}
df_risk[, high_analytical_150 := risk_analytical.x >= 1/150]
df_risk[, high_online_150 := risk_analytical.y >= 1/150]

conf_150 <- table(
  Analytical = ifelse(df_risk$high_analytical_150, "High (≥1/150)", "Low (<1/150)"),
  Online = ifelse(df_risk$high_online_150, "High (≥1/150)", "Low (<1/150)")
)

print(conf_150)

# Calculate agreement metrics
n_total_150 <- sum(conf_150)
n_agree_150 <- conf_150[1,1] + conf_150[2,2]
n_disagree_150 <- conf_150[1,2] + conf_150[2,1]

accuracy_150 <- n_agree_150 / n_total_150
sensitivity_150 <- conf_150[2,2] / (conf_150[2,2] + conf_150[1,2])
specificity_150 <- conf_150[1,1] / (conf_150[1,1] + conf_150[2,1])
kappa_150 <- (2 * (conf_150[1,1] * conf_150[2,2] - conf_150[1,2] * conf_150[2,1])) / 
             ((conf_150[1,1] + conf_150[1,2]) * (conf_150[1,2] + conf_150[2,2]) + 
              (conf_150[1,1] + conf_150[2,1]) * (conf_150[2,1] + conf_150[2,2]))

cat("\nClassification Agreement at 1/150 Threshold:\n")
cat(sprintf("  Total agreement: %.2f%% (%d/%d)\n", 
            100 * accuracy_150, n_agree_150, n_total_150))
cat(sprintf("  Disagreement: %.2f%% (%d/%d)\n", 
            100 * (1 - accuracy_150), n_disagree_150, n_total_150))
cat(sprintf("  Sensitivity: %.2f%%\n", 100 * sensitivity_150))
cat(sprintf("  Specificity: %.2f%%\n", 100 * specificity_150))
cat(sprintf("  Cohen's Kappa: %.4f\n", kappa_150))
```

```{r confusion_plot_150, fig.height=4, fig.width=6}
# Create confusion matrix plot
conf_df_150 <- as.data.frame(conf_150)
conf_df_150$prop <- conf_df_150$Freq / sum(conf_df_150$Freq)

ggplot(conf_df_150, aes(x = Online, y = Analytical, fill = Freq)) +
  geom_tile(color = "white", linewidth = 1.5) +
  geom_text(aes(label = sprintf("%d\n(%.1f%%)", Freq, 100*prop)), 
            size = 5, fontface = "bold") +
  scale_fill_gradient(low = "white", high = "steelblue") +
  labs(
    title = "Classification Agreement: 1/150 Threshold",
    x = "Online Calculator",
    y = "Analytical Formula",
    fill = "Count"
  ) +
  theme_minimal() +
  theme(
    axis.text = element_text(size = 11, face = "bold"),
    axis.title = element_text(size = 12, face = "bold"),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5)
  )
```

---

## Summary and Conclusions


---

## Session Information

```{r session_info}
sessionInfo()
```




