---
title: "Data Specifications"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data Specifications}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Data Specification

The package requires specific input data columns for risk calculation. A
demo dataset is provided to demonstrate the expected format:

```{r, eval = FALSE}
# Load the demo data
demo_data <- get_demo_data()
head(demo_data)
```

Required columns include:

-   Pregnancy details (twins, gestational age in weeks)

-   Maternal characteristics (age, height, weight, race)

-   Medical history (chronic conditions, diabetes)

-   Biophysical measurements (MAP, UtPI)

-   Biochemical measurements (PlGF, PAPP-A)

### Allowed and Disallowed Strings

Your data set must adhere to the following standards to be processed.
While the `calculate_pe_risk()` function accepts a data frame, it
converts each row to a data list using `row_to_list()`. The
`row_to_list()` function processes and standardizes specific columns.
Below are examples of allowed and disallowed values for key columns:

#### Twins Status (twins)

**Allowed:**

-   "no" → Singleton pregnancy

-   "monochorionic" → Twins (Monochorionic)

-   "dichorionic" → Twins (Dichorionic)

**Disallowed:**

-   "singleton" (use "no" instead)

-   "mono" (use "monochorionic" instead)

-   "di" (use "dichorionic" instead)

#### Smoking Status (smoking)

**Allowed:**

-   "yes" → Smoker

-   "no" → Non-smoker

**Disallowed:**

-   "y", "n" (use "yes" or "no")

-   "true", "false" (use "yes" or "no")

#### Race (race)

**Allowed (numeric codes):**

-   1 → White

-   2 → Black

-   3 → South Asian

-   4 → East Asian

-   5 → Mixed

**Disallowed:**

-   "white", "black" (use numeric codes)

#### Diabetes Drugs

**Allowed:**

-   "diet only"

-   "insulin"

-   "insulin+metformin"

-   "metformin", "Metformin"

**Disallowed:**

-   "diet"

-   "insulin + metformin" (spaces)

#### Biomarker Inclusion (include_plgf, include_pappa)

**Allowed:**

-   "no" → Not included

-   "mom", "MoM" → MoM (multiples of the median)

-   "raw", "Raw" → Raw values

**Disallowed:**

-   "none" (use "no")

-   "median", "raw data" (use "mom" or "raw")

#### Biomarker Machines (plgf_machine, pappa_machine)

**Allowed:**

-   "delfia" → Delfia machine

-   "kryptor" → Kryptor machine

-   "roche" → Roche machine

**Disallowed:**

-   "kr" (use full names)

### Validation and Conversion

To verify and convert your dataset to the appropriate format:

```{r, eval = FALSE}
# Validate dataset columns
validate_columns(demo_data)

# Convert a data row to a parameter list
data_row <- demo_data[1, ]
params <- row_to_list(data_row)
```

The `row_to_list()` function automatically handles allowed strings and
issues warnings if invalid values are detected. This ensures that data
passed to the FMF preeclampsia calculator adheres to required standards.
