---
title: "PERiskProfileR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{PERiskProfileR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
#| include: false
library("PERiskProfileR")
```

## Introduction

PERiskProfileR is an R package designed to calculate preeclampsia risk scores based on first-trimester parameters using the Fetal Medicine Foundation (FMF) algorithm. The package provides both local implementation of the risk calculations and an interface to the online FMF calculator.

## Installation

You can install PERiskProfileR from GitHub using devtools:

```r
if (!requireNamespace("devtools", quietly = TRUE))
    install.packages("devtools")
devtools::install_github("andjar/PERiskProfileR", ref = "main")
```

## Data Specification

The package requires specific input data columns for risk calculation. A demo dataset is provided to demonstrate the expected format:

```{r}
# Load the demo data
demo_data <- get_demo_data()
head(demo_data)
```

Required columns include:
- Pregnancy details (twins, gestational age in weeks)
- Maternal characteristics (age, height, weight, race)
- Medical history (chronic conditions, diabetes)
- Biophysical measurements (MAP, UtPI)
- Biochemical measurements (PlGF, PAPP-A)

### Allowed and Disallowed Strings
Your data set must adhere to the following standards to be processed. While the `calculate_pe_risk()` function accepts a data frame, it converts each row to a data list using `row_to_list()`. The `row_to_list()` function processes and standardizes specific columns. Below are examples of allowed and disallowed values for key columns:

#### Twins Status (twins)
**Allowed:**
- "no" → Singleton pregnancy
- "monochorionic" → Twins (Monochorionic)
- "dichorionic" → Twins (Dichorionic)

**Disallowed:**
- "singleton" (use "no" instead)
- "mono" (use "monochorionic" instead)
- "di" (use "dichorionic" instead)

#### Smoking Status (smoking)
**Allowed:**
- "yes" → Smoker
- "no" → Non-smoker

**Disallowed:**
- "y", "n" (use "yes" or "no")
- "true", "false" (use "yes" or "no")

#### Race (race)
**Allowed (numeric codes):**
- 1 → White
- 2 → Black
- 3 → South Asian
- 4 → East Asian
- 5 → Mixed

**Disallowed:**
- "white", "black" (use numeric codes)

#### Biomarker Inclusion (include_plgf, include_pappa)
**Allowed:**
- "no" → Not included
- "mom" → MoM (multiples of the median)
- "raw" → Raw values

Disallowed:
- "none" (use "no")
- "median", "raw data" (use "mom" or "raw")

#### Biomarker Machines (plgf_machine, pappa_machine)
**Allowed:**
- "delfia" → Delfia machine
- "kryptor" → Kryptor machine
- "roche" → Roche machine

**Disallowed:**
- "kr" (use full names)

### Validation and Conversion
To verify and convert your dataset to the appropriate format:

```{r, eval = FALSE
# Validate dataset columns
validate_columns(demo_data)

# Convert a data row to a parameter list
data_row <- demo_data[1, ]
params <- row_to_list(data_row)
```

The `row_to_list()` function automatically handles allowed strings and issues warnings if invalid values are detected. This ensures that data passed to the FMF preeclampsia calculator adheres to required standards.

## Local Risk Calculation

### Getting MoMs (Multiples of Median)

The package calculates MoM values for key biomarkers:

```{r}
# Convert a data row to parameter list
params <- row_to_list(demo_data[1,])

# Calculate MoMs for individual markers
mom_map <- get_mom_map(params)
mom_utpi <- get_mom_utpi(params)
mom_plgf <- get_mom_plgf(params)

print(c(MAP_MoM = mom_map, UtPI_MoM = mom_utpi, PlGF_MoM = mom_plgf))
```

### Getting Risk Scores

Calculate preeclampsia risk using the formula-based method:

```{r}
# Calculate risk for a single patient
risk_results <- calculate_formula_risk(params, G = 37)
print(risk_results)

# Calculate risk for multiple patients
results_df <- calculate_pe_risk(
  demo_data,
  method = "formula",
  report_as_text = FALSE,
  G = 37
)
head(results_df[, c("risk_prior", "risk")])
```

The risk scores are returned as:
- `risk_prior`: Background risk based on maternal characteristics
- `risk`: Final risk incorporating biomarker measurements

## Online Risk Calculation

The package can also submit data to the FMF online calculator:

```{r, eval=FALSE}
# Calculate risk using online calculator
online_results <- calculate_pe_risk(
  demo_data,
  method = "online",
  save_responses = TRUE,
  response_dir = "calculator_responses"
)
```

### Saving and Reading HTML Files

For documentation and retrieval, the package can save the FMF calculator response pages:

```{r, eval=FALSE}
# Extract risk scores from saved HTML files
saved_results <- extract_risk_scores_from_files(
  response_dir = "calculator_responses",
  report_as_text = FALSE
)
```

The saved HTML files contain:
- Original input data
- Calculated MoM values
- Prior and final risk scores
- Additional details about the calculations

### Comparing Online and Local Results

You can compare results from both methods:

```{r, eval=FALSE}
# Calculate risks using both methods
formula_results <- calculate_pe_risk(demo_data, method = "formula")
online_results <- calculate_pe_risk(demo_data, method = "online")

# Compare risk scores
comparison <- data.frame(
  Formula_Risk = formula_results$risk,
  Online_Risk = online_results$risk
)
print(comparison)
```

## Important Notes

1. This package is NOT intended for clinical use. All risk calculations should be verified using official clinical tools.

2. The online calculator requires internet connectivity and may be subject to rate limiting or access restrictions.

3. Always validate the results and ensure the input data meets the expected format and ranges.

## References

The package implements the algorithm described in:
- Tan MY, et al. Screening for pre-eclampsia by maternal factors and biomarkers at 11-13 weeks' gestation. Ultrasound Obstet Gynecol. 2018;52:186-195.